---
title: "Access to Improved Sanitation Facilities"
author: "Lena Nguyen"
runtime: shiny
output: 
  html_document:
    theme: "cerulean"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(httr)
library(jsonlite)
library(dplyr)
library(tidyr)
library(ggplot2)
library(DT)
```

```{r get_data, include = FALSE}
api_url <- 'http://api.worldbank.org/v2/countries/all/indicators/'
indicator <- 'SH.STA.ACSN'

# Construct URL to get specific indicator
sanit_url <- paste0(api_url, indicator)

# Get data from URL
# NOTE: Per page is set very high because there are ~15300 observations
wash <- GET(sanit_url,
            query = list(per_page = 20000, date = '1960:2018', format = 'json'))

# Turn JSON API into R object
wash_data <- fromJSON(content(wash, as = 'text'), flatten = TRUE)

# inspect data
str(wash_data)

# Take only the part of the data returned from the API with the indicator data
# Part 1 seems to only be information about number of obs and page
wash_data <- wash_data[[2]]

#--- Get basic data on country
# NOTE: Mostly interested in income data
country_url <- "http://api.worldbank.org/v2/countries/"

country <- GET(country_url,
              query = list(per_page = 1000, format = 'json'))

# Prettify nested JSON data from API
country_data <- fromJSON(content(country, "text"), flatten = TRUE)
country_data <- country_data[[2]]
```

```{r wrangling, include = FALSE}
# Look at the columns in the sanitation data
colnames(wash_data)

wash_data <- wash_data %>%
  # Select only necessary columns
  select(1:3, starts_with("country.")) %>%
  # Rename columns
  rename(iso3c = countryiso3code, year = date,
         iso2c =  country.id, country = country.value)

# Check number of missing values
sum(is.na(wash_data$value))
# Note: About 50% of the data is missing a value for the sanitation indicator

#--- Look at the character of missing data
missing <- wash_data %>%
  filter(is.na(value))

# Plot missing data
ggplot(data = missing, aes(x = year)) + geom_bar()
# Note: Looks like all data is missing except for 1990-2015
# Will just drop missing data for data for the purposes of this exercise
wash_data <- wash_data %>% filter(!is.na(value))

# Look at data structure
str(country_data)
colnames(country_data)
head(country_data)

# Keep only necessary columns and rename
# This also drops aggregated regional level data
country_data <- country_data %>%
  filter(incomeLevel.value != "Aggregates") %>%
  select(id, name, region.id, region.value, incomeLevel.id, incomeLevel.value) %>%
  rename(iso3c = id, country = name, regionID = region.id,
         region = region.value, income_code = incomeLevel.id,
         income = incomeLevel.value)

# Make data frame with the difference between 2015 and 1990 for each country
changes <- wash_data %>%
  # sort data
  arrange(country, year) %>%
  # group by country
  group_by(country) %>%
  # find difference between each row
  mutate(diff = c(NA, diff(value))) %>%
  group_by(country) %>%
  # add all the differences up for net change over the years in each country
  summarise(total_diff = round(sum(diff, na.rm = TRUE), digits = 2)) %>%
  # arrange by the difference
  arrange(total_diff) %>% 
  left_join(country_data, by = "country") %>%
  # rank countries on most improved
  mutate(rank = ifelse(iso3c != "", rank(-total_diff, ties.method = "first"), NA))

# make df with all the aggregate data
final_df <- wash_data %>% left_join(changes, by = c("country"))

regions <- c("Europe & Central Asia", "North America",
             "Latin America & Caribbean", "Middle East & North Africa",
             "South Asia", "East Asia & Pacific", "Sub-Saharan Africa",
             "World")

inc_levels <- c("Low income", "Lower middle income",
             "Upper middle income", "High income")
```

### Summary
The global trend for percentage of the population with access to improved sanitation between 1990 and 2015 is upwards increasing from 52.9% in 1990 to 67.5% in 2015. The region that showed the most improvement is East Asia Pacific with an increase of almost 24 percentage points. The three countries in the East Asia Pacific region that showed the most impressive improvements are: Palau *(53.5 percentage point increase)*, Lao PDR *(50.8 percentage point increase)*, Vietnam *(41.8 percentage point increase)*. Countries that are considered Low Income countries showed an increase of about 15 percentage points but this improvement was not equal across all regions in the world. Countries in Sub-Sarahan Africa showed an increase of only 6 percentage points between 1990 and 2015. Although there has been progress made in improving access to sanitation worldwide, there is still a lot of room for improvement because only 28% of the population in Low Income countries and only 52% of the population in Lower Middle Income countries have access to improved sanitation. A few countries saw decreases in access to improved sanitation during this time period, the majority of them being middle income countries. The three countries that saw the largest decrease are classified as Lower Middle Income. For example, Nigeria suffered from a steady decrease in the percentage of the population with access to improved sanitation from 38.1% in 1990 to 29.0% in 2015. 

### Graph and Data
```{r chart, dpi = 300, echo = FALSE}
inputPanel(
  selectInput("type", label = "Select group:",
              choices = c("Region", 
                          "Income Level", 
                          "10 Countries with Largest Increase", 
                          "10 Countries with Largest Decrease"),
              selected = "Region",
              width = "500px"),
      downloadButton("downloadData", "Download Data for Plot"),
      downloadButton("downloadPlot", "Download Plot")
)

df <- reactive({
  if (input$type == "Region"){
    final_df %>% filter(country %in% regions) %>% select(year, value, iso2c, country)
  } else if (input$type == "Income Level") {
    final_df %>% filter(country %in% inc_levels) %>% select(year, value, iso2c, country)
  } else if (input$type == "10 Countries with Largest Increase") {
    final_df %>% filter(rank <= 10)
  } else if (input$type == "10 Countries with Largest Decrease") {
    final_df %>% filter(rank >= 238)
  }
})

renderPlot({
  ggplot(data = df(),
       aes(x = year, y = value, group = country, color = country)) +
  geom_point() + geom_line() +
  scale_y_continuous(breaks = seq(10, 100, by = 5)) +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(x = "Year", y = "% of Population with Access",
       title = "Improved Sanitation Facilities (% of population with access)")
})
```

```{r table, echo = FALSE}
dt_data <- reactive({
  if (input$type == "Region") {
    changes %>% filter(country %in% regions) %>% 
      select(country, total_diff, region, income, rank) %>% arrange(rank)
  } else if (input$type == "Income Level") {
    changes %>% filter(country %in% inc_levels) %>% 
      select(country, total_diff, region, income, rank)
  } else if (input$type == "10 Countries with Largest Increase") {
    changes %>% 
      filter(iso3c != "") %>% 
      select(country, total_diff, region, income, rank) %>% 
      arrange(rank)
  } else if (input$type == "10 Countries with Largest Decrease") {
    changes %>% 
      filter(iso3c != "") %>% 
      select(country, total_diff, region, income, rank) %>% 
      arrange(desc(rank))
  }
})

renderDT(
  datatable(dt_data(),
            colnames = c("Country", "Total Change (% points)", 
                           "Region", "Income Level", "Rank"),
            rownames = FALSE
            )
)
```

```{r download, echo = FALSE}
output$downloadPlot <- downloadHandler(
    filename = function() { 
      paste("Access to Improved Sanitation by for ", 
                                  input$type, ".png", sep = "")},
    content = function(file) {
        ggsave(file, plot = last_plot(), 
               device = png(width = 10, height = 5, res = 300, units = "in"))
      }
)

output$downloadData <- downloadHandler(
    filename = function() {
      "sanitation.csv"
    },
    content = function(file) {
      write.csv(df(), file, row.names = FALSE)
})
```