---
title: "Access to Improved Sanitation"
author: "Lena Nguyen"
date: "February 22, 2018"
runtime: shiny
output: 
  html_document:
    css: my.css
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)

x <- c("httr", "jsonlite", "dplyr", "tidyr", "ggplot2", "DT")
# install.packages(x) # warning: uncommenting this may take a number of minutes
lapply(x, library, character.only = TRUE) # load the required packages
```
```{r get_data, include = FALSE}
api_url <- 'http://api.worldbank.org/v2/countries/all/indicators/'
indicator <- 'SH.STA.ACSN'

# Construct URL to get specific indicator
sanit_url <- paste0(api_url, indicator)

# Get data from URL
# NOTE: Per page is set very high because there are ~15300 observations
wash <- GET(sanit_url,
            query = list(per_page = 20000, date = '1960:2018', format = 'json'))

# Turn JSON API into R object
wash_data <- fromJSON(content(wash, as = 'text'), flatten = TRUE)

# inspect data
str(wash_data)

# Take only the part of the data returned from the API with the indicator data
# Part 1 seems to only be information about number of obs and page
wash_data <- wash_data[[2]]

#--- Get basic data on country
# NOTE: Mostly interested in income data
country_url <- "http://api.worldbank.org/v2/countries/"

country <- GET(country_url,
              query = list(per_page = 1000, format = 'json'))

# Prettify nested JSON data from API
country_data <- fromJSON(content(country, "text"), flatten = TRUE)
country_data <- country_data[[2]]
```

```{r wrangling, include = FALSE}
# Look at the columns in the sanitation data
colnames(wash_data)

wash_data <- wash_data %>%
  # Select only necessary columns
  select(1:3, starts_with("country.")) %>%
  # Rename columns
  rename(iso3c = countryiso3code, year = date,
         iso2c =  country.id, country = country.value)

# Check number of missing values
sum(is.na(wash_data$value))
# Note: About 50% of the data is missing a value for the sanitation indicator

#--- Look at the character of missing data
missing <- wash_data %>%
  filter(is.na(value))

# Plot missing data
ggplot(data = missing, aes(x = year)) + geom_bar()
# Note: Looks like all data is missing except for 1990-2015
# Will just drop missing data for data for the purposes of this exercise
wash_data <- wash_data %>% filter(!is.na(value))

# Look at data structure
str(country_data)
colnames(country_data)
head(country_data)

# Keep only necessary columns and rename
# This also drops aggregated regional level data
country_data <- country_data %>%
  filter(incomeLevel.value != "Aggregates") %>%
  select(id, name, region.id, region.value, incomeLevel.id, incomeLevel.value) %>%
  rename(iso3c = id, country = name, regionID = region.id,
         region = region.value, income_code = incomeLevel.id,
         income = incomeLevel.value)

# Make data frame with the difference between 2015 and 1990 for each country
changes <- wash_data %>%
  # sort data
  arrange(country, year) %>%
  # group by country
  group_by(country) %>%
  # find difference between each row
  mutate(diff = c(NA, diff(value))) %>%
  group_by(country) %>%
  # add all the differences up for net change over the years in each country
  summarise(total_diff = round(sum(diff, na.rm = TRUE), digits = 2)) %>%
  # arrange by the difference
  arrange(total_diff) %>% 
  left_join(country_data, by = "country") %>%
  # rank countries on most improved
  mutate(rank = ifelse(iso3c != "", rank(-total_diff, ties.method = "first"), NA))

# make df with all the aggregate data
final_df <- wash_data %>% left_join(changes, by = c("country"))

regions <- c("Europe & Central Asia", "North America",
             "Latin America & Caribbean", "Middle East & North Africa",
             "South Asia", "East Asia & Pacific", "Sub-Saharan Africa")

inc_levels <- c("Low income", "Lower middle income",
             "Upper middle income", "High income")
```

```{r, dpi = 300, echo = FALSE}
inputPanel(
  selectInput("type", label = "Select group:",
              choices = c("Region", 
                          "Income Level", 
                          "10 Countries with Largest Increase", 
                          "10 Countries with Largest Decrease"),
              selected = "Region",
              width = "500px"),
      downloadButton("downloadData", "Download Data for Plot"),
      downloadButton("downloadPlot", "Download Plot")
)

df <- reactive({
  if (input$type == "Region"){
    final_df %>% filter(country %in% regions) %>% select(year, value, iso2c, country)
  } else if (input$type == "Income Level") {
    final_df %>% filter(country %in% inc_levels) %>% select(year, value, iso2c, country)
  } else if (input$type == "10 Countries with Largest Increase") {
    final_df %>% filter(rank <= 10)
  } else if (input$type == "10 Countries with Largest Decrease") {
    final_df %>% filter(rank >= 238)
  }
})

renderPlot({
  ggplot(data = df(),
       aes(x = year, y = value, group = country, color = country)) +
  geom_point() + geom_line() +
  scale_y_continuous(breaks=seq(10, 100, by = 5)) +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(x = "Year", y = "% of Population with Access",
       title = "Improved Sanitation Facilities (% of population with access)")
})

renderDT(
  datatable(changes %>% filter(iso3c != "") %>% 
              select(country, total_diff, region, income, rank) %>% arrange(rank),
            colnames = c("Country", "Total Change (% points)", 
                           "Region", "Income Level", "Rank"),
            rownames = FALSE
            )
)

output$downloadPlot <- downloadHandler(
    filename = function() { 
      paste("Access to Improved Sanitation by for ", 
                                  input$type, ".png", sep = "")},
    content = function(file) {
        ggsave(file, plot = last_plot(), 
               device = png(width = 10, height = 5, res = 300, units = "in"))
      }
)

output$downloadData <- downloadHandler(
    filename = function() {
      "sanitation.csv"
    },
    content = function(file) {
      write.csv(df(), file, row.names = FALSE)
})